---
title: "BUA 455 Lecture 12 - Parallel Lines Models"
author: "Penelope Pooler Eisenbies"
date: "`r Sys.Date()`"
output: 
  html_document: 
    code_folding: show
    toc: yes
    toc_depth: 4
    toc_float: yes
---

### Getting Started with Markdown

- Download Zipped R project 

- Open Zipped folder and copy internal folder (R Project) to a BUA 345 folder on your computer NOT IN DOWLOADS

- Click on .Rproj file to open project

- Click on `code_data_output` file to open it.

- Click on `BUA_345_Lecture_12.Rmd` to open it.

- Run Setup Chunk.

---

### Setup

* The setup chunk shows the packages needed for this demo.   

* R will install specified packages if needed (only required once after R is installed)  

* R will load specified packaged (required every time you start a new R session)  

* The first time you run this code, R will install these packages which will be slow.  

* **If you get warnings, that's okay.**  

* If you get **error messages**, I (or TA), can help you.

---

### Setup Chunk for Lecture 12

```{r setup for Lecture 12, include = T}


# this line specifies options for default options for all R Chunks
knitr::opts_chunk$set(echo=T, highlight=T)

# suppress scientific notation
options(scipen=100)

# install helper package that loads and installs other packages, if needed
if (!require("pacman")) install.packages("pacman", repos = "http://lib.stat.cmu.edu/R/CRAN/")

# install and load required packages
pacman::p_load(pacman,tidyverse, magrittr, olsrr, gridExtra, ggiraphExtra)

# verify packages
p_loaded()

```

**NOTES:

- ** Don't worry about `xaringanthemer` package (required for my slides but not for your code).

- New Package: `ggiraphExtra`

---

### Review Question - Import data

-   Recall The Tennessee Real Estate data:

```{r import data and create transformed variable}

# import and examine data
tn_houses <- read_csv("TN_houses.csv", show_col_types = F) |>
  glimpse(width=75)

```

---

### Review Question -  Natural Log transformation

-   We will build an MLR model was using the natural log of Price (ln_Price)
-   This transformation is needed because Price is **RIGHT-SKEWED**.
-   Review of how to create this transformation

```{r ln transformation of Price}

# create new variable in tn_houses dataset
# ln_price = the natural log of price
tn_houses <- tn_houses |>
  mutate(ln_price = log(Price)) |>
  glimpse( width=75)

```

---

### Review Question - Examination Variable Distrbutions

- Histogram of `Price` (`hist_price`), shows distribution of raw data is right-skewed with high outliers.

- Histogram of `ln_price` (`hist_ln_price`) shows distribution of ln transformed data is symmetric and normally distributed.

```{r creating  and formatting histograms of Price and ln_price, fig.width=14, fig.height=7, fig.align='center'}

# histogram of original price data
hist_Price <- tn_houses |> 
  ggplot() +  
  geom_histogram(aes(x=Price), color="darkblue", fill="lightblue") + 
  labs(x="Selling Price", y="Frequency") +  
  theme_classic() + 
  theme(axis.title = element_text(size=18),
        axis.text = element_text(size=15),
        plot.background = element_rect(colour = "darkgrey", fill=NA, size=2))

# histogram of ln_price
hist_ln_price <- tn_houses |> 
  ggplot() +  
  geom_histogram(aes(x=ln_price), color="darkgreen", fill="lightgreen") + 
  labs(x="Natural Log of Selling Price", y="Frequency") +  
  theme_classic()+ 
  theme(axis.title = element_text(size=18),
        axis.text = element_text(size=15),
        plot.background = element_rect(colour = "darkgrey", fill=NA, size=2))

grid.arrange(hist_Price, hist_ln_price, ncol=2)

```

---


### Review Question -  Regression model using `ln_price`

```{r regression model using ln_price}

# saves model and prints it to screen because it is enclosed in parentheses
(tn_houses_mlr <- ols_regress(ln_price ~ Bedrooms + Bathrooms, data=tn_houses))

```

---

### Lecture 12 In-class Exercises - Back-transforming Model Estimates

#### **Question 1 (L12) - Session ID: bua345s23**

**Based on the model output, What is the estimated price of a house with 4 bedrooms and 3 bathrooms (rounded to closest $1000)?**

```{r review question incomplete R code}

# incomplete model equation
# fill in intercept and values from question
# (y_est <- ___ + 0.056*4 + 0.375*3)

# back-transform y_est using exp function
# (est_dollars <- exp(y_est))

# round to 1000
# round(est_dollars,-3)

```

NOTE: All 3 steps above could be done with one line but it is helpful to break it down when learning.

### Data Examples - R Star Wars Dataset

- Dataset of characters from Star Wars franchise

```{r eval=F}
?starwars # run this line to see help file documentation
```

```{r example from R starwars data}

my_starwars <- starwars           # save R starwars dataset to Global Environment
my_starwars |> glimpse(width=75)  # examine data

```

---

### Examining Data

-   A good way to examine categorical data variables is to examine how many observations are in each category.
-   For example, we can examine the Star Wars character data by species and gender
-   Too many species...

```{r summary of starwars species}

# summarizing starwars character data by species and gender
my_starwars |> select(species, gender) |> table()

```

---

### Lecture 12 In-class Exercises 

#### **Question 2 (L12) - Session ID: bua345s23**

**Is species a nominal or ordinal variable?**

---

### Star wars Example  - Examining categorical Data

- Since `Human` is the most common species, we can filter the data to look at those characters only.
-   Then we can examine prevalence of each gender and eye color among human characters.

```{r summary of gender and eye color}

my_starwars |> filter(species=="Human") |> # filter to humans only
  select(gender, eye_color) |>             # examine genders and eye colors
  table()

```

---

### Lecture 12 In-class Exercises

#### **Question 3 (L12) - Session ID: bua345s23**

- **Which R command is used to summarize the number of observations in each gender x eye_color combination?**

NOTE: This useful command will also be used in HW 6.

---

### Data Examples - GT cars dataset

- Deluxe automobiles from the 2014-2017 period

```{r eval=F}
?gt::gtcars # run this line to see help file documentation
```

```{r import gtcars dats}

gt_cars <- read_csv("gt_cars.csv", show_col_types = F) |> # import data
  glimpse(width=75) 

```

---

### GT Cars Example  - Examining categorical Data

```{r examine categories within gt data}

gt_cars |> select(bdy_style, year) |> table()

```
<br>

### Lecture 12 In-class Exercises

#### **Question 4 (L12) - Session ID: bua345s23**

**Which variable in gt_cars, body style (`bdy_style`) or `year` **could** be treated as ordinal?**

---

### Data Example - Celebrity Salaries Data

- Many (not all) celebrities see a decrease in their annual income as they age.

- There is a negative relationship between wages and ages.

- Is this relationship the same for males a females?

```{r import and examine celebrity data}

# import celebrity data
celeb <- read_csv("celeb.csv", show_col_types = F) |>
  glimpse(width=75) 

```

---

### Celebrity Salaries Data - Examining categories and Correlations

```{r}
celeb |> select(Gender) |> table()  # examine counts for each category

celeb |> select(Earnings, Age) |> cor() |> round(2) # examine correlation between earnings and age

```


---

### Celebrity Salaries Data - Examining Model Options

#### Model Option 1: SLR

- This model assumes no difference between males and females.
- In this case we use Base R command for regression, `lm`.
- Model created with `lm` can be used to create an interactive plot.


```{r celeb slr model}

# slr model created using lm
celeb_slr <- lm(Earnings ~ Age, data=celeb)

# create and save interactive plot
(int_clb_slr <- ggPredict(celeb_slr, interactive=T))

```

---

### Celebrity Salaries Data - Examining Model Options

#### Model Option 2: Categorical Regression Model

-   The SLR mode is okay, but we can probably do better.
-   It is logical to suspect there is a difference between males and females in terms of earnings
-   We add `Gender` to the model to test this:


```{r celeb mlr model}

# mlr model created using lm
celeb_cat_lm <- lm(Earnings ~ Age + Gender, data=celeb)

# create and save interactive plot
(int_clb_mlr <- ggPredict(celeb_cat_lm, interactive=T))

```

---

### Celebrity Salaries Data - MLR Model Output

- We can see the model equation (poorly formatted) for each gender, in the plot.
- We can also get these equations from the model output, but it requires a little work.
- Examine the model output:

```{r celeb cat mlr formal output}

# formatted regression output
# model is saved and printed to screen
(celeb_cat_ols<- ols_regress(Earnings ~ Age + Gender, data=celeb))

```

---


### Getting Model Equations from Regression Output

- By default R chooses baseline categories alphabetically

  - `Female` is before `Male` so `Female` is the baseline

  - Female SLR Model: `Est. Earnings = 134.112 - 2.37 * Age`

  - Male SLR Model:

    - `Est. Earnings = 134.112 - 2.37 * Age + 15.383`

    - `Est. Earnings = 134.112 + 15.383 - 2.37*Age`

    - `Est. Earnings = 149.505 - 2.37 * Age`

- The difference between the intercepts for Females and Males is shown in the model output.

  - Difference in intercepts is labeled with name of categorical variable and category

  - Difference (Increase) for Males is labeled **`GenderMale`** and equals **`15.383`**

---

### Lecture 12 In-class Exercises

#### **Question 5 (L12) - Session ID: bua345s23**

**Based on our categorical regression model, is the difference between male and female earnings (approx. 15 $M), statistically significant?**

HINT:  Look at the p-value for the **`GenderMale`** term in the model to answer this question.

---

### Data Example - House Remodeling Data

-   What is the effect of remodeling on house selling price?

```{r import and examine house remodel data}

# import house remodeling data
house_remodel <- read_csv("house_remodel.csv", show_col_types = F) |>
  glimpse(width=75) 

```

---

### House Remodeling Data - Examine Categories and Correlations

-   What is the effect of remodeling on house selling price?

```{r  examine categories and correlations} 

# examine number of obs in each category
house_remodel |> select(Remodeled) |> table() 

# correlation between price and square feet
house_remodel |> select(Price, Square_Feet) |> cor() |> round(2)

```


---

### House Remodeling Data - Examining Model Options

#### Model Option 1: SLR

- SLR model assumes no difference due to remodeling
- Again, we use Base R command for regression, `lm`
- Model created with `lm` can be used to create an interactive plot.


```{r house remodel slr model}

# slr model created using lm
rem_slr <- lm(Price ~ Square_Feet, data=house_remodel)

# create and save interactive plot
(int_rem_slr <- ggPredict(rem_slr, interactive=T))

```

---

### House Remodeling Data - Examining Model Options

#### Model Option 2: Categorical Regression Model

- SLR mode is okay, but there is likely to be a difference between `Remodeled` and un-`Remodeled` houses.
- To test for that difference we add the categorical variable `Remodeled` to the model:

```{r house remodel mlr model}

# mlr model created using lm
rem_cat_lm <- lm(Price ~ Square_Feet + Remodeled, data=house_remodel)

# create and save interactive plot
(int_rem_mlr <- ggPredict(rem_cat_lm, interactive=T))

```

---

### House Remodel dataa - MLR Model Output

- We can see the model equation (poorly formatted) for each category, in the plot.
- We can also get these equations from the model output, but it requires a little work.
- Examine the model output:

```{r house rem cat mlr formal output}

# formatted regression output
# model is saved and printed to screen
(house_rem_ols<- ols_regress(Price ~ Square_Feet + Remodeled, data=house_remodel))

```

---

### Getting Model Equations from Regression Output

- By default R chooses baseline categories alphabetically

  - `No` is before `Yes` so un-`Remodeled` houses are the baseline

  - un-Remodeled SLR Model: `Est. Price = 137549.093 + 137.879 * Square_Feet`

  - Remodeled SLR Model:

    - `Est. Price = 137549.093 + 137.879 * Square_Feet + 90917.216`

    - `Est. Price = 137549.093 + 90917.216 + 137.879 * Square_Feet`

    - `Est. Price = _____ + 137.879 * Square_Feet`

---

### Lecture 12 In-class Exercises

#### **Question 6 (L12) - Session ID: bua345s23**

The difference between the intercepts for `Remodeled` and un-`Remodeled` houses is shown in the model output.

- Difference in intercepts is labeled with name of categorical variable and category

- Difference (Increase) for remodeling is labeled **`RemodeledYes`** and equals **`90917.216`**

**What is the **intercept** for the prices of 'Remodeled' houses in the Categorical Regression model (Round to closest thousand ($K).**

---

#### **Question 7 (L12) - Session ID: bua345s23**

**Based on our categorical regression model, is the difference in selling price between remodeled (Remodeled = `Yes`) and un-remodeled (Remodeled = `No`) homes statistically significant?**

---

### **Key Points from Today**

- Categorical Parallel Lines Model

  - Separate SLR model for each category.
  - Modeling categories simultaneously with one mode is
    -   more efficient
    -   more accurate 

-   Next Similar Model BUT each category has a different slope
 
-   HW 6 is posted (Due 3/6)
 
-   First set of data is **almost** identical to `house_remodel` data.

<br>

You may submit an 'Engagement Question' about each lecture until midnight on the day of the lecture. **A minimum of four submissions are required during the semester.**


